---
title: "Statistical_Analysis_Project"
output:
  pdf_document:
    latex_engine: xelatex

date: "2025-09-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Data Load
```{r}
# Load required packages
library(readr)
library(dplyr)
library(ggplot2)

# Read the CSV file
df <- read_csv("Statistical Analysis/Project/IBM_HR_Employee.csv")

# View the structure of the data
str(df)

# See first few rows
head(df)
```

# Data Cleaning
```{r}
# Remove rows with any missing values
df_clean <- df %>% na.omit()

# Check for duplicates
df_clean <- df_clean %>% distinct()

# Check unique values for key variables
unique(df_clean$Attrition)
unique(df_clean$OverTime)
unique(df_clean$Department)
```


# Save the cleaned dataset as a new CSV file
```{r}
write.csv(df_clean, "Statistical Analysis/Project/IBM_HR_Employee_Cleaned.csv", row.names = FALSE)
```


# EDA
```{r}
# View dataset dimensions
dim(df_clean)

# Quick summary of all variables
summary(df_clean)

# See the first 10 rows
head(df_clean, 10)

# Check variable names
colnames(df_clean)
```

# Categorical Variables
```{r}
# Identify categorical variable names
cat_vars <- c("Attrition", "OverTime", "Department", "Gender", "EducationField", "JobRole",
              "MaritalStatus", "BusinessTravel")


df_clean[cat_vars] <- lapply(df_clean[cat_vars], as.factor)

# Check all the selected are categorical variables
# str(df_clean[cat_vars])

```


# Check for Missing Values
```{r}
colSums(is.na(df_clean))
```



# Module 0 - Descriptive Statistics Report

The dataset used in this project is the IBM HR Employee Attrition dataset. It contains information on 1,470 employees, 
with 35 variables including demographic information, job roles, compensation, work environment, and attrition status.
A full variable list is included in the appendix. In this report, we focus on a selected group of variables that are most 
relevant for both business analysis and the statistical concepts covered in this course.

```{r}
# Select key variables for Module 0
df_selected <- df %>%
  select(Attrition, Department, OverTime, Age, MonthlyIncome, YearsAtCompany)

# View the structure
str(df_selected)
summary(df_selected)
head(df_selected)
```

## Categorical Variables
```{r}
# Attrition
table(df_clean$Attrition)
prop.table(table(df_clean$Attrition))
```


```{r}
# Department
table(df_clean$Department)
prop.table(table(df_clean$Department))
```


```{r}
# OverTime
table(df_clean$OverTime)
prop.table(table(df_clean$OverTime))
```

## Numerical Variables

```{r}
# Age, MonthlyIncome, YearsAtCompany

summary(df_clean$Age)
summary(df_clean$MonthlyIncome)
summary(df_clean$YearsAtCompany)

```



## Visualization

```{r}
library(ggplot2)
library(dplyr)

# Calculate mean and median age
mean_age <- mean(df_clean$Age)
median_age <- median(df_clean$Age)

ggplot(df_clean, aes(x = Age)) +
  geom_histogram(bins = 20, fill = "#56B4E9", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean_age), color = "red", linetype = "dashed", size = 1, show.legend = TRUE) +
  geom_vline(aes(xintercept = median_age), color = "darkgreen", linetype = "dotted", size = 1, show.legend = TRUE) +
  labs(
    title = "Age Distribution of Employees",
    subtitle = paste0("Mean Age: ", round(mean_age, 1), "; Median Age: ", median_age),
    x = "Employee Age",
    y = "Number of Employees",
    caption = "Red dashed = Mean | Green dotted = Median"
  ) +
  theme_minimal(base_size = 14)

```

```{r}
# Compute quantiles
q1 <- quantile(df_clean$MonthlyIncome, 0.25)
med <- quantile(df_clean$MonthlyIncome, 0.5)
q3 <- quantile(df_clean$MonthlyIncome, 0.75)
```


```{r}
ggplot(df_clean, aes(x = "", y = MonthlyIncome)) +
  geom_boxplot(fill = "gold", color = "black", width = 0.4, outlier.color = "red", outlier.shape = 16) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 3, fill = "blue", color = "black") +
  
  annotate("text", x = 1.13, y = q1, label = paste0("Q1: $", q1), hjust = 0, color = "black", size = 4) +
  annotate("text", x = 1.13, y = med, label = paste0("Median: $", med), hjust = 0, color = "blue", size = 4) +
  annotate("text", x = 1.13, y = q3, label = paste0("Q3: $", q3), hjust = 0, color = "black", size = 4) +
  labs(
    title = "Boxplot of Monthly Income with Quantiles",
    subtitle = paste0("Q1: $", q1, " | Median: $", med, " | Q3: $", q3),
    y = "Monthly Income (USD)",
    x = NULL,
    caption = "Red dots = outliers; Blue diamond = median"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```




```{r}
# Relationships between Variables

table(df_clean$Attrition, df_clean$Department)
prop.table(table(df_clean$Attrition, df_clean$Department), 2)

```



# Module 1 - Probability Analysis

Let’s use the variables: Attrition (Yes/No), OverTime (Yes/No).

## Attrition vs OverTime

## Joint Probability

```{r}
# Create contingency table
joint_table <- table(df_selected$Attrition, df_selected$OverTime)
joint_table

# Joint probabilities
joint_prob <- prop.table(joint_table)
joint_prob
```



## Conditional Probability
```{r}
# Conditional probability: P(Attrition | OverTime)
cond_prob_attrition_given_overtime <- prop.table(joint_table, 2)
cond_prob_attrition_given_overtime

# Conditional probability: P(OverTime | Attrition)
cond_prob_overtime_given_attrition <- prop.table(joint_table, 1)
cond_prob_overtime_given_attrition
```


## Bayes’ Theorem

Let’s compute: P(Attrition=Yes | OverTime=Yes) using Bayes’ theorem.

Formula: 

```{r}
# Calculate P(OverTime="Yes" | Attrition="Yes")
p_overtime_yes_given_attrition_yes <- joint_table["Yes", "Yes"] / sum(joint_table["Yes", ])

# Calculate P(Attrition="Yes")
p_attrition_yes <- sum(df_selected$Attrition == "Yes") / nrow(df_selected)

# Calculate P(OverTime="Yes")
p_overtime_yes <- sum(df_selected$OverTime == "Yes") / nrow(df_selected)

# Bayes' theorem
p_attrition_yes_given_overtime_yes <- (p_overtime_yes_given_attrition_yes * p_attrition_yes) / p_overtime_yes
p_attrition_yes_given_overtime_yes

```



### Attrition vs. Department
Try more variables

```{r}
# 1. Create contingency table
attr_dept_table <- table(df_selected$Attrition, df_selected$Department)
attr_dept_table

# 2. Joint probability table
attr_dept_joint_prob <- prop.table(attr_dept_table)
attr_dept_joint_prob

# 3. Conditional probability: P(Attrition | Department)
# (Probability of attrition status within each department)
cond_prob_attr_given_dept <- prop.table(attr_dept_table, 2)
cond_prob_attr_given_dept

# 4. Conditional probability: P(Department | Attrition)
# (Probability of department within each attrition group)
cond_prob_dept_given_attr <- prop.table(attr_dept_table, 1)
cond_prob_dept_given_attr

```

## Bayes’ Theorem

P(Attrition = Yes | Department = Sales) using Bayes’ theorem:

```{r}

# P(Department = Sales | Attrition = Yes)
p_dept_sales_given_attr_yes <- attr_dept_table["Yes", "Sales"] / sum(attr_dept_table["Yes", ])
# P(Attrition = Yes)
p_attr_yes <- sum(df_selected$Attrition == "Yes") / nrow(df_selected)
# P(Department = Sales)
p_dept_sales <- sum(df_selected$Department == "Sales") / nrow(df_selected)

# Apply Bayes’ theorem
p_attr_yes_given_dept_sales <- (p_dept_sales_given_attr_yes * p_attr_yes) / p_dept_sales
p_attr_yes_given_dept_sales

```


# Module 2 - Random Variables, PMF, Expectation, Variance, and CDF

##  Define the Random Variable - Discrete

Let X = "YearsAtCompany" for a randomly selected employee, which is a discrete random variable 
(showing how many years an employee has worked at the company).

```{r}

# Get the random variable
X <- df_selected$YearsAtCompany

# PMF: Frequency table of YearsAtCompany
pmf_table <- table(X) / length(X)
pmf_table

# Convert to a data frame for plotting
pmf_df <- as.data.frame(pmf_table)
colnames(pmf_df) <- c("YearsAtCompany", "Probability")


# Plot the PMF
library(ggplot2)

# Find the max
max_point <- pmf_df[which.max(pmf_df$Probability), ]

ggplot(pmf_df, aes(x = as.numeric(as.character(YearsAtCompany)), y = Probability)) +
  geom_col(fill = "steelblue", width = 0.7) +
  
  geom_text(data = max_point, 
            aes(label = paste0("Max: ", round(Probability, 3))), 
            vjust = -0.5, size = 3.5, color = "red") +
  
  labs(
    title = "Probability Mass Function (PMF) of Employee Years at the Company",
    subtitle = paste0("The most common year is ", max_point$YearsAtCompany, 
                      " years with an estimated probability of (P = ", round(max_point$Probability, 3), ")"),
    x = "Years at Company",
    y = "Probability"
  ) +
  theme_minimal(base_size = 11.5)

```

## Expectation (Mean) and Variance
```{r}
# Expectation (mean)
mean_X <- mean(X)
mean_X

# Variance
var_X <- var(X)
var_X

# Standard Deviation
sd_X <- sd(X)
sd_X
```



## Cumulative Distribution Function (CDF)

```{r}
# Empirical CDF
cdf_X <- ecdf(X)

# Plot the CDF
# plot(cdf_X, main="Empirical CDF of Years at Company", xlab="Years at Company", ylab="F(x)")

# What proportion of employees have been at the company ≤ 5 years?
cdf_X(5)
```
```{r}
# Empirical CDF
cdf_X <- ecdf(X)

# Compute key probability: P(X ≤ 5)
p_le5 <- cdf_X(5)

# Plot the CDF
plot(cdf_X, 
     main = "Empirical CDF of Years at Company",
     xlab = "Years at Company", 
     ylab = "F(x)", 
     col = "darkblue", lwd = 2, pch = 16, cex = 0.8)

# Highlight the point at X = 5
points(5, p_le5, col = "red", pch = 19, cex = 1.5)
text(5, p_le5, labels = paste0(" P(X ≤ 5) = ", round(p_le5, 3)),
     pos = 4, col = "red", cex = 0.9, font = 2)

# Add grid for readability
grid()

```



##  Define the Random Variable - Continuous

Let Y = "MonthlyIncome" for a randomly selected employee, which is a continuous random variable,
showing the monthly income of employees.

### Probability Density Function (PDF)

```{r}
density_income <- density(df_selected$MonthlyIncome)
head(density_income$x)
head(density_income$y)

```


### Visualize the Distribution of variable MonthlyIncome
```{r}
# Histogram
ggplot(df_selected, aes(x = MonthlyIncome)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Monthly Income", x = "Monthly Income", y = "Count")
```

```{r}
# Density Plot

library(ggplot2)

# Density estimation
dens <- density(df_selected$MonthlyIncome)
max_idx <- which.max(dens$y)
max_x <- dens$x[max_idx]
max_y <- dens$y[max_idx]

# Plot
ggplot(df_selected, aes(x = MonthlyIncome)) +
  geom_density(fill = "orange", alpha = 0.6, color = "black", size = 1) +
  geom_vline(xintercept = max_x, linetype = "dashed", color = "red", size = 1) +
  annotate("point", x = max_x, y = max_y, color = "red", size = 3) +
  annotate("text", x = max_x, y = max_y, 
         label = paste0("Peak ≈ $", round(max_x,0)), 
         vjust = 2, color = "darkblue", fontface = "bold", size = 5) +
  labs(
    title = "Probability Density Function (PDF) of Monthly Income",
    subtitle = paste0("Most common income is around $", round(max_x,0),
                      " (highest density)"),
    x = "Monthly Income (USD)",
    y = "Density"
  ) +
  theme_minimal(base_size = 12)
```


## Calculate Expectation (Mean) and Variance

```{r}
# Mean (expectation)
mean_income <- mean(df_selected$MonthlyIncome)
mean_income

# Variance
var_income <- var(df_selected$MonthlyIncome)
var_income

# Standard deviation
sd_income <- sd(df_selected$MonthlyIncome)
sd_income

```




## Empirical Cumulative Distribution Function (CDF)
```{r}
# Empirical CDF
cdf_income <- ecdf(df_selected$MonthlyIncome)

# Plot the CDF
# plot(cdf_income, main = "Empirical CDF of Monthly Income", xlab = "Monthly Income", ylab = "F(x)")

# Example: Proportion of employees earning <= $5000
cdf_income(5000)

```

```{r}
# Empirical CDF
cdf_income <- ecdf(df_selected$MonthlyIncome)

p_le_5000 <- cdf_income(5000)                 # P(X ≤ 5000)
median_inc <- median(df_selected$MonthlyIncome)

# Plot
plot(cdf_income,
     main = "Empirical CDF of Monthly Income",
     xlab = "Monthly Income (USD)", ylab = "F(x)",
     col = "black", lwd = 2, pch = 16, cex = 0.8)


# Show P(X ≤ 5000)
abline(v = 5000,           col = "red",  lty = 2)
abline(h = p_le_5000,      col = "red",  lty = 2)
points(5000, p_le_5000,    col = "red",  pch = 19, cex = 1.3)
text(5000, p_le_5000, 
     labels = paste0(" P(X ≤ 5000) = ", round(p_le_5000, 3)),
     pos = 4, col = "red", cex = 0.95, font = 2)

abline(v = median_inc, col = "blue", lty = 3)
points(median_inc, 0.5, col = "blue", pch = 19, cex = 1.2) # 标记点（可选）
text(median_inc, 0.52, 
     labels = paste0("Median ≈ $", round(median_inc, 0)),
     pos = 3, col = "blue", cex = 0.9, font = 2)


grid(col = "gray90")

```




## Test for Normality

```{r}
# QQ Plot
qqnorm(df_selected$MonthlyIncome)
qqline(df_selected$MonthlyIncome, col = "red")
```


