---
title: "Module Project 2"
output: pdf_document
date: "2025-09-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r}
# Load the dataset (the cleaned version from Module Project 1)

df <- read.csv("Statistical Analysis/Project/IBM_HR_Employee_Cleaned.csv")

# View summary
summary(df$MonthlyIncome)
```


Define the random variable for your data using one of the probability distributions covered in the course:
- MonthlyIncome as a continuous random variable (modeled using Normal distribution)

```{r}
# Check summary of MonthlyIncome

summary(df$MonthlyIncome)

```


```{r}
# The value of mean, standard deviation, variance and skewness

income_mean <- mean(df$MonthlyIncome)
income_sd <- sd(df$MonthlyIncome)
income_var <- var(df$MonthlyIncome)
income_skew <- skewness(df$MonthlyIncome)


# Show the results of mean and variance

cat("Mean Monthly Income:", income_mean, "\n")
cat("Standard Deviation of Monthly Income:", income_sd, "\n")
cat("Variance of Monthly Income:", income_var, "\n")
cat("Skewness of Monthly Income:", income_skew, "\n")
```


```{r}
library(e1071)
library(scales)

# Plot
ggplot(df, aes(x = MonthlyIncome)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 40, 
                 fill = "skyblue", 
                 color = "black", 
                 alpha = 0.7) +
  geom_density(color = "red", size = 1.2) +
  geom_vline(xintercept = income_mean, color = "blue", linetype = "dashed", size = 1) +
  annotate("text", x = income_mean, y = 0.00022, 
           label = paste0("Mean = $", format(round(income_mean), big.mark = ",")), 
           color = "blue", vjust = -0.5, size = 4) +
  labs(
    title = "Distribution of Monthly Income",
    x = "Monthly Income (USD)",
    y = "Density",
    caption = paste0(
      "Red line represents the estimated density curve\n",
      "Mean = $", format(round(income_mean, 1), big.mark = ","), 
      " | SD = ", round(income_sd, 1), 
      " | Variance = ", round(income_var, 1), 
      " | Skewness = ", round(income_skew, 2)
    )
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.y = element_text(margin = margin(r = 10))
  )


```



```{r}
# Calculate conditional expectations

# Summarize the mean MonthlyIncome and standard deviation for each JobLevel

conditional_mean <- df %>%
  group_by(JobLevel) %>%
  summarise(
    Count = n(),
    MeanIncome = mean(MonthlyIncome),
    SDIncome = sd(MonthlyIncome)
  )

print(conditional_mean)
```

Explanation:
Each bar shows the average MonthlyIncome (expected value) for a specific JobLevel. 
As JobLevel increases, we expect MonthlyIncome to increase, which helps understand how income is structured across career levels in the organization.



```{r}
# Plot: Bar chart with error bars and value labels

ggplot(conditional_mean, aes(x = factor(JobLevel), y = MeanIncome)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = MeanIncome - SDIncome, ymax = MeanIncome + SDIncome),
                width = 0.2, color = "black") +
  
  geom_text(aes(label = format(round(MeanIncome), big.mark = ",")), 
            vjust = -0.5, hjust = 1.2, size = 3.5, fontface = "bold", color = "red") +

  geom_text(aes(y = MeanIncome + SDIncome + 300, 
                label = paste0("±SD=$", format(round(SDIncome), big.mark = ","))), 
            hjust = -0.2, size = 3, color = "blue") +
  labs(
    title = "Expected Monthly Income by Job Level",
    x = "Job Level",
    y = "Average Monthly Income (USD)",
    caption = "Error bars represent ±1 standard deviation"
  ) +
  theme_minimal() +
  ylim(0, max(conditional_mean$MeanIncome + conditional_mean$SDIncome) + 2000)

```


```{r}
# Calculate covariance and correlations
# We'll analyze the relationship between MonthlyIncome, Age, and YearsAtCompany

# Covariance
cov_income_age <- cov(df$MonthlyIncome, df$Age)
cov_income_years <- cov(df$MonthlyIncome, df$YearsAtCompany)

```

```{r}
# Correlation
cor_income_age <- cor(df$MonthlyIncome, df$Age)
cor_income_years <- cor(df$MonthlyIncome, df$YearsAtCompany)
```


```{r}
# Show the results of covariance and correlation

cat("Covariance (Monthly Income, Age):", cov_income_age, "\n")
cat("Covariance (Monthly Income, YearsAtCompany):", cov_income_years, "\n")

cat("Correlation (Monthly Income, Age):", cor_income_age, "\n")
cat("Correlation (Monthly Income, YearsAtCompany):", cor_income_years, "\n")
```


```{r}
# We'll analyze correlations between: MonthlyIncome, Age, YearsAtCompany

library(corrplot)

# Select relevant variables
corr_vars <- df %>% select(MonthlyIncome, Age, YearsAtCompany)

# Compute correlation matrix
cor_matrix <- cor(corr_vars, use = "complete.obs")

# Print correlation matrix
print(cor_matrix)

```

```{r}
# Plot
# install.packages("ggcorrplot")
library(ggcorrplot)

# Choose Variables - Age, YearsAtCompany and MonthlyIncome
selected_vars <- df[, c("Age", "YearsAtCompany", "MonthlyIncome")]

# Calculate the matrix
cor_matrix <- cor(selected_vars, use = "complete.obs")

# Plot
ggcorrplot(cor_matrix,
           method = "square",
           type = "lower",
           lab = TRUE,
           lab_size = 4,
           lab_col = "black",
           show.diag = TRUE,
           colors = c("red", "white", "steelblue"),
           outline.color = "gray",
           title = "Correlation Matrix of Key Quantitative Variables") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_blank()
  )

```

